Human Factors Influencing Visual Statistical Inference

### 2015-05-17 sociology_paper.tex
footnote for calculus

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examines these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect on performance. There is evidence that a learning effect exists but only in that observers get faster with repeated evaluations, but not more often correct. The placement of data plot in the lineup does not affect the inference.

@@

The lineup protocol introduced in \citet{buja:2009} can be used to test the significance of \hh{graphical} findings in the exploratory data analysis process. This methodology is  part of what is called visual statistical inference, which has been developed further  and validated by \citet{majumder:2013} in simulation studies of head to head comparisons with conventional inference. One of the major contributions of  \citet{majumder:2013}  is to define the power of the visual test and provide methods to obtain the power for a particular lineup. It was observed that the power can be as good or better than that of a conventional test in some scenarios.

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are showing  data generated under a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 \hh{out} of 72 observers picked the  plot \hh{of the observed data} (located in panel number $3^2+4$ in the lineup).  The corresponding $p$-value is \hh{0.0077}, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

>>>>

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while  exploring data with graphics.  Visual inference relies on human observers inspecting a lineup: one  data plot randomly placed among a set of decoy plots. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way of estimating the effectiveness of a statistical test is its power. The estimate of power of a lineup is controlled by combining evaluations from multiple observers. Factors that may  affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examines these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect on performance. There is evidence that a learning effect exists but only in the sense that observers become faster with repeated evaluations, while accuracy stays the same. The placement of data plot in the lineup does not affect the inference.

@@

The lineup protocol introduced in \citet{buja:2009} can be used to quantify the significance of \hh{graphical} findings in the exploratory data analysis process. This methodology is  part of what is called visual statistical inference, which has been developed further  and validated by \citet{majumder:2013} in simulation studies of head to head comparisons with conventional inference. One of the major contributions of  \citet{majumder:2013}  is to define the power of the visual test and provide methods to obtain the power for a particular lineup. It was observed that the power can be as good or better than that of a conventional test in some scenarios.

@@

Figure~\ref{fig:elect-1} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are showing  data generated under a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 \hh{out} of 72 observers picked the  plot \hh{of the observed data}, located in panel number $3^2+4$ in the lineup \footnote{The little piece of calculus used for describing the location of the data plot in the lineup provides a small cognitive hurdle that is supposed to enable the interested reader to inspect the lineup without being biased by knowing the location of the data plot.}.  The corresponding $p$-value is \hh{0.0077}, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).
